---
title: "Illustration of RW model and its RPE--varying learning rate"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = FALSE)
```

```{r}
library(data.table)
library(tidyverse)
library(ggplot2)

theme_set(theme_bw(base_size=12))
clPalette <- c('#999999','#E69F00', '#56B4E9', "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

# Setting for simulation

### RW model
We consider RW model
\begin{align*}
V_{t+1} &= V_t + \alpha \delta_t \\
\delta_t &= r_t - V_t
\end{align*}
where $V_1 = 0$. 

### fMRI signal model

* Model 1 (value)
\begin{align*}
y_t = \beta_{V} V_t + \epsilon_t
\end{align*}

* Model 2 (RPE)
\begin{align*}
y_t = \beta_{\delta} \delta_t + \epsilon_t
\end{align*}

$\epsilon_t$ is Gaussian variable with mean zero, sd, $\sigma_{\epsilon} = 0.5$, 
$\beta_{\delta} = 1$, $\beta_{V} = 1$

### Setting

* 100 trials
* Reward probability: $p_r = 0.4$
* true parameters: $\alpha = 0.5$, $\rho = 1.0$
* fit parameters: $\hat{\alpha} = 0.2, 0.5, 0.8$ for Group 1,2,3. 
$\hat{\rho} = 1.0$ for all groups 

# Run simulation

```{r}
# number of trials
nTrial <- 100

# seed for random numbers
set.seed(1)

# reward probability
pr <- 0.4

# dataframe for data
df <- data.table()

dfstat <- data.table()
idxGroup <- 0

# sd of fMRI noise
sigma_epsion <- 0.1

# true alpha (common to both groups)
rho.true <- 1.0

# true alpha (common to both groups)
alpha.true <- 0.5

# fit parmaeters for each group
fit.rho.array <- c(1, 1, 1) # reward sensitivity for group 1, group 2, group 3
fit.alpha.array <- c(0.2, alpha.true, 0.8) # learning rate for group 1, group 2, group 3

# reward sequence
r <- c(rep(1, round(nTrial * pr)), rep(0, round(nTrial * (1-pr))))
r <- sample(r)

dfseq <- data.table()

# true model
delta <- numeric(nTrial) 
V <- numeric(nTrial)

for (t in 1:nTrial) {
  delta[t] <- rho.true * r[t] - V[t]
  
  if (t < nTrial) {
    V[t+1] <- V[t] + alpha.true * delta[t]
  }
}

# generate fMRI signal
y.delta<- delta + rnorm(nTrial, mean = 0, sd = sigma_epsion) 
y.delta<- y.delta - mean(y.delta) # mean-centered 

y.V <- V + rnorm(nTrial, mean = 0, sd = sigma_epsion) 
y.V <- y.V - mean(y.V) # mean-centered 

for (idxGroup in seq_along(fit.rho.array)) {
  
  rho.fit <- fit.rho.array[[idxGroup]]
  alpha.fit <- fit.alpha.array[[idxGroup]]
  
  # initialize value 
  V.fit <- numeric(nTrial)
  
  delta.fit <- numeric(nTrial) 
  
  # fit model
  for (t in 1:nTrial) {
    
    delta.fit[t] <- rho.fit * r[t] - V.fit[t]
    
    if (t < nTrial) {
      V.fit[t+1] <- V.fit[t] + alpha.fit * delta.fit[t]
    }
  }
  
  dftmp <- data.table(group = as.factor(idxGroup)
                      ,trial = 1:nTrial
                      ,r = r
                      ,delta = delta - mean(delta)
                      ,delta.fit = delta.fit - mean(delta.fit)
                      ,y.delta= y.delta
                      ,y.V = y.V
                      ,V = V
                      ,V.fit = V.fit)
  
  
  glm.delta <- glm("y.delta~ delta.fit", data = dftmp)
  
  glm.V <- glm("y.V ~ V.fit", data = dftmp)
  
  dfseq <- rbind(dfseq, dftmp)
  
  dftmp2 <- data.table(group = as.factor(idxGroup),
                       # RL model parameters ---- 
                       alpha.fit = alpha.fit, 
                       alpha.true = alpha.true,
                       rho.fit = rho.fit,
                       alpha.true = alpha.true,
                       # latent variable statistics ---- 
                       cor.delta.fit.y = cor(delta.fit, y.delta),
                       cov.delta.fit.y = cov(delta.fit, y.delta),
                       cov.V.fit.y = cov(V.fit, y.V),
                       cor.V.fit.y = cor(V.fit, y.V),
                       var.delta.fit = var(delta.fit), 
                       var.V.fit = var(V.fit),
                       # regression ---- 
                       beta.delta = glm.delta$coefficients["delta.fit"],
                       beta.V = glm.V$coefficients["V.fit"]
  )
  dfstat <- dplyr::bind_rows(dfstat, dftmp2)
}
```

# Plot results

```{r}
theme_set(theme_classic(base_size = 14))
```

## Temporal sequence

```{r}
gV <- ggplot(dfseq, aes(x = trial, group = group, color = group)) + 
  geom_point(aes(y = r), color = "black") +
  geom_line(aes(y = V.fit, size = group)) +
  geom_point(aes(y = V.fit, size = group, shape=group),size = 1.5) +
  scale_size_manual(values = c(1.2,1.1,1.0)) + 
  scale_color_manual(values = clPalette) +
  xlim(0,40) + 
  xlab("Trial number") + ylab("Value, Reward")

gRPE <- ggplot(dfseq, aes(x = trial, group = group, color = group)) +
  geom_line(aes(y = delta.fit, size = group)) +
  geom_point(aes(y = delta.fit, size = group, shape=group),size = 1.5) +
  scale_size_manual(values = c(1.2,1.1,1.0)) + 
  scale_color_manual(values = clPalette) +
  xlim(0,40) + 
  xlab("Trial number") + ylab("RPE")

library(gridExtra)
gridExtra::grid.arrange(gV, gRPE, nrow = 2)

gg <- arrangeGrob(gV, gRPE, nrow = 2)
ggsave(file="./figs/illustration_sequence_alpha.eps", gg)
```

## RPE 

```{r}
g1 <- mutate(dfseq, group = factor(group)) %>%
  ggplot(aes(x = delta.fit, y = y.delta, color = group, shape=group)) +
  geom_point(size = 2) + 
  geom_smooth(method = lm, se=FALSE) + theme(legend.position = "none") +
  scale_color_manual(values = clPalette) + 
  theme(aspect.ratio=1)
print(g1)
```


```{r}
dfstat %>% select(cor.delta.fit.y, cov.delta.fit.y, var.delta.fit, beta.delta) %>% 
  round(digits = 3)
```

## Value

```{r}
g2 <- mutate(dfseq, group = factor(group)) %>%
  ggplot(aes(x = V.fit, y = y.V, color = group, shape=group)) +
  geom_point(size = 2) + 
  geom_smooth(method = lm, se=FALSE) +
  scale_color_manual(values = clPalette) +
   theme(legend.position = "none") +
  theme(aspect.ratio=1)
print(g2)

print(dfstat)
```

```{r}
gg <- arrangeGrob(g2, g1, nrow = 1)
ggsave(file="./figs/illustration_correlation_alpha.eps", gg)
```

```{r}
dfstat %>% select(cor.V.fit.y, var.V.fit, beta.V) %>% 
  round(digits = 3)
```



