---
title: "Simulation_forgetting_EXCUTE (N=30, Reverse = 1)"
author: "toyama"
date: "2020.09.15"
output:
  html_document:
    theme: default
    toc: true
    toc_depth: 2
---



```{r}
source("Simulation_forgetting_FUNCTION.R")
library(scales)
```


# Simulation Procedure

1. Data generation
  - `func_datageneration(Model, trueparam, RHtype, N)`: Execute this!
  - `func_qlearning (Model, param, Est, c, r, Sim, r1, r2)`
  : Includeing in the above function with Sim = 1
2. Parameter Fit
  - `func_ParameterFit(Est, Model, DF)`: Execute this!
  - `func_minimize(Model, param, Est, c, r)`: included in the above function
  - `func_qlearning (Model, param, Est, c, r, Sim, r1, r2)`
  : Includeing in the above function with Sim = 0 which is defined in the above function

# Setting

## Setting: Task

```{r Setting_task}
set.seed(1)
nTrial = 180
trial = seq(1,nTrial,1)
pr <- c(0.8,0.2)
RHtype <- 1 # Reverse Time
```


## Setting: True parameters
* High-F group: $\beta = 4$, $\alpha = 0.5$, $\phi = 0.4$
* Low-F group : $\beta = 4$, $\alpha = 0.5$, $\phi = 0.05$
* $\mu$ (defautl value parameter) : 0.5 (The same $\mu$ is set in the func_qlearning function.)

```{r Setting_trueparameters}
mu <- 0.5
trueparam_HighF<-c(4,0.5,0.4) # corresponds to "beta","alpha","phi" 
trueparam_LowF<-c(4,0.5,0.05) 
```


# Data generation & Fitting & Regression & Effect size
N: 30 each
```{r}
set.seed(1)
N = 30
RHtype <- 1 # num of reverse. num of blocks is RHtype + 1

Model = "BLFH"
trueparam = trueparam_HighF
rbetter <- c(rep(1,pr[1]*nTrial/(1+RHtype)),rep(0,pr[2]*nTrial/(1+RHtype)))
rworse <- c(rep(1,pr[2]*nTrial/(1+RHtype)),rep(0,pr[1]*nTrial/(1+RHtype)))
func_datageneration(Model, trueparam, RHtype, N)
DF_FH <-DF %>% tbl_df()



Model = "BLFL"
trueparam = trueparam_LowF
rbetter <- c(rep(1,pr[1]*nTrial/(1+RHtype)),rep(0,pr[2]*nTrial/(1+RHtype)))
rworse <- c(rep(1,pr[2]*nTrial/(1+RHtype)),rep(0,pr[1]*nTrial/(1+RHtype)))
func_datageneration(Model, trueparam, RHtype, N)
DF_FL <-DF %>% tbl_df()

########
# Fitting
#######

#----------------------------- Fit DF_FH by noForget model---------------------------#
Est <- "ML"
Data = DF_FH
Model = "BL"
func_ParameterFit(Est,Model,Data)
FHnoF_param <- DF_param %>% mutate(group = "FHnoF") 
FHnoF <- dplyr::left_join(Data, DF_fit, by=c("id","trial")) %>% mutate(group = "FHnoF")

#----------------------------- Fit DF_FL by noForget model---------------------------#
Est <- "ML"
Data = DF_FL
Model = "BL"
func_ParameterFit(Est,Model,Data)
FLnoF_param <- DF_param %>% mutate(group = "FLnoF")
FLnoF <- dplyr::left_join(Data, DF_fit, by=c("id","trial")) %>% mutate(group = "FLnoF")

#----------------------------- Fit DF_FH by Forget model---------------------------#
Est <- "ML"
Data = DF_FH
Model = "BLF"
func_ParameterFit(Est,Model,Data)
FHF_param <- DF_param %>% mutate(group = "FHF")
FHF <- dplyr::left_join(Data, DF_fit, by=c("id","trial")) %>% mutate(group = "FHF")

#----------------------------- Fit DF_FL by Forget model---------------------------#
Est <- "ML"
Data = DF_FL
Model = "BLF"
func_ParameterFit(Est,Model,Data)
FLF_param <- DF_param %>%  mutate(group = "FLF")
FLF <- dplyr::left_join(Data, DF_fit, by=c("id","trial")) %>%  mutate(group = "FLF")

All <- dplyr::bind_rows(FHnoF,FHF,FLnoF,FLF)
Allparam <- dplyr::bind_rows(FHnoF_param,FHF_param,FLnoF_param,FLF_param) %>%
  select(-starts_with("NA")) %>% 
  tbl_df() %>% 
  mutate(beta = as.numeric(beta),
         alphaL = as.numeric(alphaL),
         alphaF = as.numeric(alphaF)
  )


########
# Regression
#######


func_Regression_Reverse(All)
Reg_All <- temp_Reg_All


########
# Effect size
#######

func_cohenD(Reg_All)
cohenD_All <-  temp_cohenD_All


```

## t test function

```{r}
ttest <- function(x1,x2,pair){
  t <- t.test(x1,x2, paired=pair,var.equal = T)
  tvalue <- round(t$statistic,2)
  df <- t$parameter
  pvalue <- round(t$p.value,4)
  cohenD <- round(cohen.d(x1,x2, paired=pair)$estimate,4)
  #cat(paste("t(",df,") = ",tvalue, " (p = ", pvalue, "; d = ", cohenD,")", sep=""))
  label_d <<- sprintf("d = %.2f", cohenD)
  #label_p <<- scales::pvalue(t$p.value, add_p = TRUE)
  label_p <<- scales::pvalue(t$p.value, prefix = c("p < ","p = ","p > "))
  return(list(df=df,tvalue=tvalue,pvalue=pvalue,cohenD=cohenD))

}
```


## t test & Plot
```{r Regression_plot}

theme_set(theme_classic(base_size=14))

str_Beta <- c("beta.GLM1.delta",
              "beta.GLM2.r",
              "beta.GLM2.nQ")

str_y <- c("beta value (RPE)",
           "beta value (r)",
           "beta value (-V)")

str_title <- c("GLM1 RPE(R-V)",
               "GLM2 reward(r)",
               "GLM2 value(-V)")

#--------
# (A) Fit without forgetting 
#--------

for (i in 1:3){
  Reg_All_noF <- Reg_All %>% filter(group %in% c("FHnoF","FLnoF"))
  y <- Reg_All %>% filter(group == "FHnoF") %>% select(str_Beta[i])
  x <- Reg_All %>% filter(group == "FLnoF") %>% select(str_Beta[i])
  ttest(x[[1]],y[[1]],0)
  
  g <- ggplot(Reg_All_noF, aes_string(x="group", y=str_Beta[i])) + 
    stat_summary(fun=mean,geom="bar",fill=c("indianred","steelblue"),
                 size=2, width=0.8) + 
    geom_jitter(width = 0.1, shape = 1) + 
    stat_summary(fun.data=mean_se, geom="errorbar", 
                 size=1,width=0.3) +
    xlab("")+  ylab(str_y[i]) +  ggtitle(str_title[i])+
    scale_x_discrete(labels=c("HighF", "LowF"))+
    scale_y_continuous(breaks=seq(-0.5,2,0.5), limits = c(-0.2,2.0))+
    geom_hline(yintercept = 0, colour = "gray44", linetype ="dashed") +
    annotate("text", x = 1.5, y = 1.9, label = label_d, size = 5) +
    annotate("text", x = 1.5, y = 1.65, label = label_p, size = 5) +
    theme(plot.title = element_text(hjust=0.5,size=16))
  
  if(i == 1){
    g <- g+labs(tag = "A")
  } else {
    g <- g+labs(tag = " ")
  }
  assign(paste("A", i, sep=""), g)
}

#--------
# (B) Fit With forgetting 
#--------

for (i in 1:3){
  Reg_All_F <- Reg_All %>% filter(group %in% c("FHF","FLF"))
  y <- Reg_All %>% filter(group == "FHF") %>% select(str_Beta[i])
  x <- Reg_All %>% filter(group == "FLF") %>% select(str_Beta[i])
  ttest(x[[1]],y[[1]],0)
  
  g <- ggplot(Reg_All_F, aes_string(x="group", y=str_Beta[i])) + 
    stat_summary(fun=mean,geom="bar",fill=c("indianred","steelblue"),
                 size=2, width=0.8) + 
    geom_jitter(width = 0.1, shape = 1) + 
    stat_summary(fun.data=mean_se, geom="errorbar", 
                 size=1,width=0.3) +
    xlab("")+  ylab(str_y[i]) +  ggtitle(str_title[i])+
    scale_x_discrete(labels=c("HighF", "LowF"))+
    scale_y_continuous(breaks=seq(-0.5,2,0.5), limits = c(-0.2,2.0))+
    geom_hline(yintercept = 0, colour = "gray44", linetype ="dashed") +
    annotate("text", x = 1.5, y = 1.9, label = label_d, size = 5) +
    annotate("text", x = 1.5, y = 1.65, label = label_p, size = 5) +
    theme(plot.title = element_text(hjust=0.5,size=16))
  
  if(i == 1){
    g <- g+labs(tag = "B")
  }else{
    g <- g+labs(tag = " ")
  }
  assign(paste("B", i, sep=""), g)
}

g <- grid.arrange(A1,A2,A3,B1,B2,B3,nrow=2)

#ggsave("Figure5.jpg", width = 10, height = 10, plot = g)
ggsave(file="./figs/group_comparison_forgetting.eps", width = 10, height = 10, plot = g)
```

## Correlation between hypothetical neural signals and calculated RPEs.

```{r Regression_correlation}
All %>%
  group_by(id,group) %>% 
  summarise(cor = cor(y,RPE_fit)) %>% 
  tbl_df() %>% 
  group_by(group) %>% 
  summarise(M_cor = mean(cor),
            SD_cor = sd(cor))
```


