---
title: "Simulation of group comparison of regression coefficients for RPE in RW model"
author: "Kentaro Katahira"
date: "2020.09.15"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = FALSE)
```

```{r}
library(data.table)
library(dplyr)
library(pwr)
library(gridExtra)
library(ggplot2)
```

fit learning rate: 0.4 -> 0.3

# Setting

### RW model
We consider RW model
\begin{align*}
V_{t+1} &= V_t + \alpha \delta_t \\
\delta_t &= r_t - V_t
\end{align*}
where $V_1 = 0$. 

### GLM

* GLM1
\begin{align*}
y_t = \beta_{\delta} \delta_t + \epsilon_t
\end{align*}

* GLM2
\begin{align*}
y_t =  \beta_{r} r_t + \beta_{nV} (-V) + \epsilon_t
\end{align*}

* GLM2'(GLM2p)
\begin{align*}
y_t = \beta_{r} r_t + \beta_{\delta} \delta_t + \epsilon_t
\end{align*}


* Model for fMRI signal: in GLMs, 
$\epsilon_t$ is Gaussian variable with mean zero, sd, $\sigma_{\epsilon} = 0.5$, 
$\beta_{\delta} = 1$, $\beta_{r} = 0$. 

### Simulation setting

* 100 trials per subject
* Reward probabilities is 0.4
* true parameter
  * Group 1 ($N=20$, depressed): $\alpha = 0.2$
  * Group 2 ($N=20$, control): $\alpha = 0.4$
* fit parameter $\hat{\alpha} = 0.3$. 


## Define function for calculating effect size (Cohen's d)

```{r}
effsize <- function(x1, x2) {
  list(x12 = ( mean(x2) - mean(x1) ) / sqrt( (var(x1) + var(x2))/2 ),
       x1 = mean(x1) / sqrt((var(x1))),
       x2 = mean(x2) / sqrt((var(x2)))
  )
}
```


# Run simulation

```{r}
# number of subjects for each group
nSub <- 20

# number of trials
nTrial <- 100

# seed for random numbers
set.seed(1)

# reward probability
pr <- 0.4

# dataframe for data
df <- data.table()

dfstat <- data.table()
idxGroup <- 0

# sd of fMRI noise
sigma_epsion <- 0.5

# true parameter
alpha1 <- 0.3
alpha2 <- 0.5

# fit alpha (common to both groups)
alpha.fit <- 0.4

# reward sequence
r <- c(rep(1, round(nTrial * pr)), rep(0, round(nTrial * (1-pr))))
r <- sample(r)
# all subjects experience the same reward sequence as in Li & Daw (2011) 

for (alpha.true in c(alpha1,alpha2)) {
  idxGroup <- idxGroup + 1
  
  for (idxSub in 1:nSub) {

    if (idxSub %% 100 == 1)
      cat("idxSub:", idxSub, format(Sys.time(), "%H:%M:%OS"), "\n")
    
    # true model
    delta <- numeric(nTrial) 
    V <- numeric(nTrial)
    
    # fit model
    V.fit <- numeric(nTrial)
    delta.fit <- numeric(nTrial) 
    
    # simulate single session
    for (t in 1:nTrial) {
      
      # true model
      delta[t] <- r[t] - V[t]
      
      if (t < nTrial) {
        V[t+1] <- V[t] + alpha.true * delta[t]
      }
      
      # fit model
      delta.fit[t] <- r[t] - V.fit[t]
      
      if (t < nTrial) {
        V.fit[t+1] <- V.fit[t] + alpha.fit * delta.fit[t]
      }
    }
    
    # generate fMRI signal
    y <- delta + rnorm(nTrial, mean = 0, sd = sigma_epsion) 
    y <- y - mean(y) # mean-centered 
    
    # record trials data
    dftmp <- data.table(group = as.factor(idxGroup),
                        subject = as.factor(idxSub),
                        trial = 1:nTrial,
                        r = r - mean(r),
                        delta = delta - mean(delta),
                        delta.fit = delta.fit - mean(delta.fit),
                        y = y, 
                        V = V,
                        nV.fit = -V.fit - mean(-V.fit))
    
    fit.GLM1 <- glm("y ~ delta.fit", data = dftmp)
    fit.GLM2 <- glm("y ~ r + nV.fit", data = dftmp)
    fit.GLM2p <- glm("y ~ r + delta.fit", data = dftmp)
    
    # record GLM results
    dftmp2 <- data.table(group = as.factor(idxGroup),
                         subject = idxSub,
                         # RL model parameters ---- #
                         alpha.fit = alpha.fit, 
                         alpha.true = alpha.true,
                         # beta values (regression coefficients) ---- #
                         beta.GLM1.delta = fit.GLM1$coefficients["delta.fit"],
                         beta.GLM2.r = fit.GLM2$coefficients["r"],
                         beta.GLM2.nV = fit.GLM2$coefficients["nV.fit"], 
                         beta.GLM2p.r = fit.GLM2p$coefficients["r"],
                         beta.GLM2p.delta = fit.GLM2p$coefficients["delta.fit"]
                         )
    
    dfstat <- dplyr::bind_rows(dfstat, dftmp2)
  }
}
```


#  Group comparison of regression coefficients for RPE 

Define function for group comparison (effect size, t-test, plotting beta values) 

```{r function for plot}
theme_set(theme_classic(base_size=15))

group_comparison <- function(df,
                      xvar = "group",
                      yvar = "beta.GLM1.delta",
                      title = "GLM1, RPE (r - V)",
                      ylabstr = "beta value",
                      tag = ""
                      ) {
  
  # t-test, calc effect size ------------
  
  data1 <- df[df$group == 1, ..yvar][[1]]
  data2 <- df[df$group == 2, ..yvar][[1]]
  
  cohen_d <- effsize(data1,data2)$x12
  t.result <- t.test(data1, data2, paired = F, var.equal = T)
  print(t.result)
  
  # plot --------------------------------
  
  label_d <- sprintf("d = %.2f", cohen_d)
  label_p <- scales::pvalue(t.result$p.value,add_p = TRUE) 
  
  g <- ggplot(df, aes_string(x = xvar, y = yvar, fill = xvar)) + 
    stat_summary(fun = mean, geom = "bar", size = 2, width = 0.8) + 
    geom_jitter(width = 0.1, shape = 1, size = 2) + 
    stat_summary(fun.data = mean_se, geom="errorbar", size=1,width=0.3) + 
    scale_fill_manual(values =c("indianred","steelblue")) + 
    xlab("Group") + 
    geom_hline(yintercept = 0, colour = "gray44", linetype ="dashed") +
    theme(axis.text.x = element_text(size=16),
          axis.text.y = element_text(size=16)) + 
    theme(legend.position = "none") + ylim(-1,2) +
    annotate("text", x = 1.5, y = 1.9, label = label_d, size = 5) + 
    annotate("text", x = 1.5, y = 1.65, label = label_p, size = 5) + 
    scale_x_discrete(breaks = c('1', '2'),
                     labels = c("Group 1", "Group 2")) +
    labs( title = title,
          x = NULL,
          y = "beta value", tag = tag)
  
  return(list(g = g, cohen_d = cohen_d, t.result = t.result))
}
```

## GLM1

```{r}
res <- group_comparison(dfstat, 
                        yvar = "beta.GLM1.delta", 
                        tag = "A")
g.GLM1 <- res$g
```

```{r, fig.height=4.5, fig.width=4}
gridExtra::grid.arrange(g.GLM1, nrow = 1)
```



## GLM2

### reward

```{r}
res <- group_comparison(dfstat, yvar = "beta.GLM2.r", 
                 title = "GLM2, reward (r)",
                 tag = "B")
g.GLM2.r <- res$g
```


### -V

```{r}
res <- group_comparison(dfstat, yvar = "beta.GLM2.nV", 
                 title = "GLM2, negative value (-V)",
                 tag = "")
g.GLM2.nV <- res$g
```



```{r, fig.height=4.5, fig.width=8}
gridExtra::grid.arrange(g.GLM2.r, g.GLM2.nV, nrow = 1)
```



## GLM2p


### reward

```{r}
res <- group_comparison(dfstat, yvar = "beta.GLM2p.r", 
                 title = "GLM2', reward (r)",
                 tag = "C")
g.GLM2p.r <- res$g
```


### -V

```{r}
res <- group_comparison(dfstat, yvar = "beta.GLM2p.delta", 
                 title = "GLM2', RPE (r - V)",
                 tag = "")
g.GLM2p.delta <- res$g
```


```{r, fig.height=4.5, fig.width=8}
gridExtra::grid.arrange(g.GLM2p.r, g.GLM2p.delta, nrow = 1)
```

```{r, fig.height=13.5, fig.width=8}
require(grid)
dummy.text <- textGrob("True value: \nalpha = 0.2 (Group 1)\n alpha = 0.4 (Group 2)\nFit value: \nalpha = 0.3")


gridExtra::grid.arrange(g.GLM1, dummy.text, 
                        g.GLM2.r, g.GLM2.nV, 
                        g.GLM2p.r, g.GLM2p.delta, 
                        nrow = 3)

gg <- arrangeGrob(g.GLM1, dummy.text, 
                        g.GLM2.r, g.GLM2.nV, 
                        g.GLM2p.r, g.GLM2p.delta, 
                        nrow = 3)
ggsave(file="./figs/group_comparison_RW_alpha.eps", gg)
```

