---
title: "Effect of within-group individual differences"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = TRUE)
```

```{r}
library(data.table)
library(tidyverse)
library(pwr)

clPalette <- c("steelblue", "indianred")
```

We examine the effect of within-group variance in true learning rates on group comparison of regression coefficients of RPE. 

# Setting 

### RW model
We consider RW model
\begin{align*}
V_{t+1} &= V_t + \alpha \delta_t \\
\delta_t &= r_t - V_t
\end{align*}
where $V_1 = 0$. 

### Model for fMRI signal: in GLMs, 
$\epsilon_t$ is Gaussian variable with mean zero, sd, $\sigma_{\epsilon} = 0.5$, 
$\beta_{\delta} = 1$, $\beta_{r} = 0$. 

### Simulation setting

* 100 trials per subject
* Reward probabilities is 0.4
* true parameter
  $\alpha$ were drawn from a (truncated) Gaussian 
distributions whose mean is 0.2 for Group 1 and 0.4 for Group 2,and s.d. are v
aried as 0.05-0.5 with the step size 0.05. 
Sample values of less than 0 were replaced with
0.0, and sample values larger than 1 were replaced with 1.0. 
* fit parameter $\hat{\alpha} = 0.3$. 
 
To obtain stable estimates of the effect size, we simulated $n = 5000$ hypothetical subjects for each group. 

# Simulation

## Define a function for calculating effect size of group difference. 

```{r}
effsize <- function(x1, x2) {
  list(x12 = ( mean(x2) - mean(x1) ) / sqrt( (var(x1) + var(x2) )/2 ),
       x1 = mean(x1) / sqrt((var(x1))),
       x2 = mean(x2) / sqrt((var(x2)))
  )
}
```

## Run simulation

```{r}
# number of subjects for each group
nSub <- 5000 

# number of trials
nTrial <- 100

# seed for random numbers
set.seed(1)

# reward probability
pr <- 0.4

# dataframe for data
df <- data.table()

# sd of fMRI noise
sigma_epsion <- 0.5

# means of true parameter for each group
alpha1 <- 0.2
alpha2 <- 0.4

# fit alpha (common to both groups)
alpha.fit <- 0.3

# reward sequence
r <- c(rep(1, round(nTrial * pr)), rep(0, round(nTrial * (1-pr))))
r <- sample(r)

df_effect_size <- data.table()

alpha.sigma.list <- c(seq(0.05, 0.5, by = 0.05))

for (alpha.sigma in alpha.sigma.list) {
  
  idxGroup <- 0
  dfstat <- data.table()
  
  for (alpha.truemean in c(alpha1,alpha2)) {
    
    idxGroup <- idxGroup + 1
    
    for (idxSub in 1:nSub) {
      
      alpha.true <- min(max(alpha.truemean + rnorm(1,0,alpha.sigma), 0), 1)
      
      if (idxSub %% 1000 == 1)
        cat("idxSub:", idxSub, format(Sys.time(), "%H:%M:%OS"), "\n")
      
      # true model
      delta <- numeric(nTrial) 
      V <- numeric(nTrial)
      
      for (t in 1:nTrial) {
        delta[t] <- r[t] - V[t]
        
        if (t < nTrial) {
          V[t+1] <- V[t] + alpha.true * delta[t]
        }
      }
      
      # initialize value 
      V.fit <- numeric(nTrial)
      
      delta.fit <- numeric(nTrial) 
      
      # fit model
      for (t in 1:nTrial) {
        
        delta.fit[t] <- r[t] - V.fit[t]
        
        if (t < nTrial) {
          V.fit[t+1] <- V.fit[t] + alpha.fit * delta.fit[t]
        }
      }
      
      # fMRI signal
      y <- delta + rnorm(nTrial, mean = 0, sd = sigma_epsion) 
      y <- y - mean(y) # mean-centered 
      
      dftmp <- data.table(group = as.factor(idxGroup),
                          subject = as.factor(idxSub),
                          trial = 1:nTrial,
                          r = r,
                          delta = delta - mean(delta),
                          delta.fit = delta.fit - mean(delta.fit),
                          y = y, 
                          V = V,
                          nV.fit = -V.fit - mean(-V.fit))
      
      fit.GLM1 <- glm("y ~ delta.fit", data = dftmp)
      fit.GLM2 <- glm("y ~ r + nV.fit", data = dftmp)
      
      dftmp2 <- data.table(
        group = as.factor(idxGroup),
        subject = idxSub,
        # RL model parameters ---- #
        alpha.fit = alpha.fit, 
        alpha.truemean = alpha.truemean,
        alpha.true = alpha.true,
        # regression ---- #
        beta.GLM1.delta = fit.GLM1$coefficients["delta.fit"],
        beta.GLM2.r = fit.GLM2$coefficients["r"],
        beta.GLM2.nV = fit.GLM2$coefficients["nV.fit"]
      )
      dfstat <- dplyr::bind_rows(dfstat, dftmp2)
    }
  }
  
  es.GLM2.nV <- effsize(dfstat$beta.GLM2.nV[dfstat$group == 1],
                dfstat$beta.GLM2.nV[dfstat$group == 2])
  
  es.GLM1.delta <- effsize(dfstat$beta.GLM1.delta[dfstat$group == 1],
                dfstat$beta.GLM1.delta[dfstat$group == 2])
  
  es.alpha <- effsize(dfstat$alpha.true[dfstat$group == 1],
                dfstat$alpha.true[dfstat$group == 2])
  
  sd.alpha <- mean(sd(dfstat$alpha.true[dfstat$group == 1]),
                sd(dfstat$alpha.true[dfstat$group == 2]))

  df_effect_size <- dplyr::bind_rows(
    df_effect_size, 
    data.frame(alpha.sigma = alpha.sigma, 
               sd.alpha = sd.alpha,
               es.GLM2.nV = es.GLM2.nV$x12, 
               es.GLM1.delta = es.GLM1.delta$x12, 
               es.alpha = es.alpha$x12))
}
```


## Plot effect sizes

Plot effect size of group differences as a function of effect size of learning rate

```{r}
theme_set(theme_classic(base_size = 14))

g <- ggplot(df_effect_size, aes(x = es.alpha , y=es.GLM2.nV)) + 
  geom_line(color = clPalette[[1]]) + 
  geom_point(color = clPalette[[1]]) + 
  geom_line(aes(y = es.GLM1.delta), color = clPalette[[2]]) + 
  geom_point(aes(y = es.GLM1.delta), color = clPalette[[2]]) + 
  xlab("effect size of alpha") + 
  ylab("effect size (between-group, RPE)") +
  theme(axis.text.x = element_text(size=15),
        axis.text.y = element_text(size=15)) + 
    theme(legend.position = "none") + ylim(0,2)
print(g)
```

Plot effect size of group differences as a function of s.d. of learning rate

```{r}
g <- ggplot(df_effect_size, aes(x=alpha.sigma, y=es.GLM2.nV)) + 
  geom_line(color = clPalette[[1]], size=1.2) + 
  geom_point(shape = 0,size = 3, color = clPalette[[1]]) + 
  geom_line(aes(y = es.GLM1.delta), size=1.2, color = clPalette[[2]]) + 
  geom_point(shape = 1,size = 3, aes(y = es.GLM1.delta), color = clPalette[[2]]) + 
  geom_line(aes(y = es.alpha), size=1.2, color = "gray40", linetype = "longdash") + 
  geom_point(shape = 2,size = 3, aes(y = es.alpha), color = "gray40") + 
  xlab("alpha.sigma") + 
  ylab("effect size (between-group, RPE)") +
  theme(axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18)) + ylim(0,5)

print(g)
ggsave(file="./figs/within_group_variance_alpha_sigma.eps", g)
```


Plot effect size of group differences as a function of s.d. of learning rate

```{r}
g <- ggplot(df_effect_size, aes(x=sd.alpha, y=es.GLM2.nV)) + 
  geom_line(color = clPalette[[1]], size=1.2) + 
  geom_point(shape = 0,size = 3, color = clPalette[[1]]) + 
  geom_line(aes(y = es.GLM1.delta), size=1.2, color = clPalette[[2]]) + 
  geom_point(shape = 1,size = 3, aes(y = es.GLM1.delta), color = clPalette[[2]]) + 
  geom_line(aes(y = es.alpha), size=1.2, color = "gray40", linetype = "longdash") + 
  geom_point(shape = 2,size = 3, aes(y = es.alpha), color = "gray40") + 
  xlab("s.d. of learning rate") + 
  ylab("effect size (between-group)") +
  theme(axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18)) + ylim(0,5)

print(g)
ggsave(file="./figs/within_group_variance_alpha_sd.eps", g)
```
