---
title: "Theory (analytical results) for group comparison of regression coefficients for RW model: dependence on number of trials"
author: "Kentaro Katahira"
date: "2020.09.18"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = FALSE)
```

```{r}
library(data.table)
library(tidyverse)
library(gridExtra)
library(pwr)
library(ggthemes)
```

# Setting

### RW model
We consider RW model
\begin{align*}
V_{t+1} &= V_t + \alpha \delta_t \\
\delta_t &= r_t - V_t
\end{align*}
where $V_1 = 0$. 

### GLMs

* GLM1
\begin{align*}
y_t = \beta_{\delta} \delta_t + \epsilon_t
\end{align*}

* GLM2
\begin{align*}
y_t =  \beta_{r} r_t + \beta_{nV} (-V) + \epsilon_t
\end{align*}


### Setting

* various number of trials ($T = 1,...,500$)
* Reward probability: $p_r = 0.4$
* Group 1 ($N=20$, depressed): $\alpha_1 = 0.2$
* Group 2 ($N=20$, control): $\alpha_2 = 0.4$
* fit parameter $\hat{\alpha} = 0.3$. 


```{r}
# number of subjects for each group
nSub <- 20

# number of trials
nTrial <- 100

# reward probability
pr <- 0.4

# sd of fMRI noise
sigma_epsilon <- 0.5

# true parameter
alpha1 <- 0.2
alpha2 <- 0.4

# fit alpha (common to both groups)
alpha.fit <- 0.3

nTrial.list <- c(1,500)
alpha.diff.list <- c(0.05, 0.1,0.15)
```

# Calculate statistical power for difference in group mean


## GLM1

```{r}
beta.GLM1.delta.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.delta = 1, rho.fit = 1, rho.true = 1) {
   (2 - alpha.fit) * (alpha.true + alpha.fit)/
  (2 * (alpha.fit + alpha.true - alpha.fit * alpha.true) ) * rho.true / rho.fit 
}

beta.GLM1.delta.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 * (2-alpha.fit) / (2 * nTrial * pr * (1 - pr) * rho.fit^2)
}
```


## GLM2

Function for analytical results. 

```{r}
beta.GLM2.nV.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
   alpha.true * (2-alpha.fit) /
    (alpha.true + alpha.fit - alpha.true * alpha.fit) * beta.nV * rho.true / rho.fit 
}

beta.GLM2.nV.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 * (2 - alpha.fit) / (nTrial * alpha.fit * pr * (1 - pr) * rho.fit^2)
}

beta.GLM2.r.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
    beta.r * rho.true
}

beta.GLM2.r.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 / (nTrial * pr * (1 - pr))
}
```


# Analytical Cohen's d 

## GLM1, RPE

```{r}
cohen.d.theory.GLM1 <- function(nTrial, 
                                pr, 
                                alpha.fit, 
                                alpha1, 
                                alpha2,
                                sigma_epsilon = 0.5) {
  # group 1
  M1.GLM1 <- beta.GLM1.delta.mean(nTrial = nTrial, pr = pr, 
                                  alpha.fit = alpha.fit, 
                                  alpha.true = alpha1)
  
  V1.GLM1 <- beta.GLM1.delta.var(nTrial, pr, 
                                 alpha.fit = alpha.fit, 
                                 alpha.true = alpha1, 
                                 sigma_epsilon = sigma_epsilon)
  
  # group 2
  M2.GLM1 <- beta.GLM1.delta.mean(nTrial, pr, 
                                  alpha.fit = alpha.fit, 
                                  alpha.true = alpha2)
  
  V2.GLM1 <- beta.GLM1.delta.var(nTrial, pr, 
                                 alpha.fit = alpha.fit, 
                                 alpha.true = alpha2, 
                                 sigma_epsilon = sigma_epsilon)
  
  # cohen_d.GLM1 
  (M2.GLM1 - M1.GLM1)/(sqrt((V1.GLM1 + V2.GLM1)/2))
  
}
```

```{r}
pwr.50 <- pwr.t.test(n = 20, sig.level = 0.05, d = NULL,  power = 0.5, 
                           type="t",alternative="two.sided")
pwr.50

pwr.80 <- pwr.t.test(n = 20, sig.level = 0.05, d = NULL,  power = 0.8, 
                           type="t",alternative="two.sided")
pwr.80
```


## GLM2.nV.

```{r}
cohen.d.theory.GLM2.nV <- function(nTrial, 
                                pr, 
                                alpha.fit, 
                                alpha1, 
                                alpha2,
                                sigma_epsilon = 0.5) {
  # group 1
  M1.GLM2.nV <- beta.GLM2.nV.mean(nTrial = nTrial, pr = pr, 
                                  alpha.fit = alpha.fit, 
                                  alpha.true = alpha1)
  
  V1.GLM2.nV <- beta.GLM2.nV.var(nTrial, pr, 
                                 alpha.fit = alpha.fit, 
                                 alpha.true = alpha1, 
                                 sigma_epsilon = sigma_epsilon)
  
  # group 2
  M2.GLM2.nV <- beta.GLM2.nV.mean(nTrial, pr, 
                                  alpha.fit = alpha.fit, 
                                  alpha.true = alpha2)
  
  V2.GLM2.nV <- beta.GLM2.nV.var(nTrial, pr, 
                                 alpha.fit = alpha.fit, 
                                 alpha.true = alpha2, 
                                 sigma_epsilon = sigma_epsilon)
  
  #cohen_d.GLM2.nV
  (M2.GLM2.nV - M1.GLM2.nV)/(sqrt((V1.GLM2.nV + V2.GLM2.nV)/2))
}
```

## plot setting

```{r}
require(latex2exp)
theme_set(theme_classic(base_size = 16))

nTrial.list <- c(1,500)
alpha.diff.list <- c(0.05, 0.1, 0.2)

gcommon <- ggplot(data.frame(nTrial = nTrial.list), aes(x = nTrial)) +
  geom_hline(yintercept = pwr.50$d, colour = "gray24", linetype ="dashed") +
  geom_hline(yintercept = pwr.80$d, colour = "gray24", linetype ="dashed") +
  labs(color=TeX('$\\alpha_2 - \\alpha_1$')) +
  scale_colour_ptol() + 
  xlab("Number of trials") + 
  ylab("effect size (between-group)")  +
  ylim(0, 3.8) + 
  theme(plot.title = element_text(hjust=0.5,size=18)) 
```

plot result of GLM1 

```{r}
g.m.GLM1 <- gcommon + 
  mapply(
    function(alpha.diff, co)
      stat_function(fun=cohen.d.theory.GLM1, 
                    args = list(
                      pr = pr,  
                      alpha.fit = alpha.fit,
                      alpha1 = alpha.fit - alpha.diff/2, 
                      alpha2 = alpha.fit + alpha.diff/2
                    ),
                    size = 1.5, 
                    aes_q(color=co)
      ), alpha.diff.list, as.factor(alpha.diff.list)) + 
  labs(tag = "A") + 
  ggtitle("GLM1, RPE") 
print(g.m.GLM1) 
```




plot 

```{r}
g.m.GLM2 <- gcommon + 
  mapply(
    function(alpha.diff, co)
      stat_function(fun=cohen.d.theory.GLM2.nV, 
                    args = list(# nTrial = nTrial, 
                      pr = pr, # sigma_epsilon = sigma_epsilon, 
                      alpha.fit = alpha.fit,
                      alpha1 = alpha.fit - alpha.diff/2, 
                      alpha2 = alpha.fit + alpha.diff/2
                      #sigma_epsilon = 0.5
                    ),
                    size = 1.5, 
                    aes_q(color=co)
      ), alpha.diff.list, as.factor(alpha.diff.list)) + 
  labs(tag = "B") + 
  ggtitle("GLM2, -V")
print(g.m.GLM2) 
```


```{r, fig.height=4.5, fig.width=12}
gridExtra::grid.arrange(g.m.GLM1, g.m.GLM2, nrow = 1)
 
gg <- arrangeGrob(g.m.GLM1, g.m.GLM2, nrow = 1)
ggsave(file="./figs/between_group_effectsize_nTrial.eps", gg)
```

# Simulation for Cohen's d

## Define function for calculating effect size (Cohen's d)

```{r}
effsize <- function(x1, x2) {
  list(x12 = ( mean(x2) - mean(x1) ) / sqrt( (var(x1) + var(x2))/2 ),
       x1 = mean(x1) / sqrt((var(x1))),
       x2 = mean(x1) / sqrt((var(x1)))
  )
}
```


## Simulation setting

```{r}
# seed for random numbers
set.seed(1)

# dataframe for data
df <- data.table() # for simulation data 

df_effectsize <- data.table() # for effect size

dfstat <- data.table()
idxGroup <- 0

# number of simulation
nEpisode <- 200
```


## define function for each subject
```{r}
run_each_sub <- function(alpha.true, alpha.fit, r, sigma_epsion) {
  
  # initialize values 
  nTrial <- length(r)
  
  # true model
  delta <- numeric(nTrial) 
  V <- numeric(nTrial)
  
  # fit model
  V.fit <- numeric(nTrial)
  delta.fit <- numeric(nTrial) 
  
  # simulate single session
  for (t in 1:nTrial) {
    
    # true model
    delta[t] <- r[t] - V[t]
    
    if (t < nTrial) {
      V[t+1] <- V[t] + alpha.true * delta[t]
    }
    
    # fit model
    delta.fit[t] <- r[t] - V.fit[t]
    
    if (t < nTrial) {
      V.fit[t+1] <- V.fit[t] + alpha.fit * delta.fit[t]
    }
  }
  
  # generate fMRI signal
  y <- delta + rnorm(nTrial, mean = 0, sd = sigma_epsion) 
  y <- y - mean(y) # mean-centered 
  
  # record trials data
  dftmp <- data.table(group = as.factor(idxGroup),
                      subject = as.factor(idxSub),
                      trial = 1:nTrial,
                      r = r,
                      delta = delta - mean(delta),
                      delta.fit = delta.fit - mean(delta.fit),
                      y = y, 
                      V = V,
                      nV.fit = -V.fit - mean(-V.fit))
  
  fit.GLM1 <- glm("y ~ delta.fit", data = dftmp)
  fit.GLM2 <- glm("y ~ r + nV.fit", data = dftmp)
  fit.GLM2p <- glm("y ~ r + delta.fit", data = dftmp)
  
  # record GLM results
  dftmp2 <- data.table(
                       # RL model parameters ---- #
                       alpha.fit = alpha.fit, 
                       alpha.true = alpha.true,
                       # regression ---- #
                       beta.GLM1.delta = fit.GLM1$coefficients["delta.fit"],
                       beta.GLM2.r = fit.GLM2$coefficients["r"],
                       beta.GLM2.nV = fit.GLM2$coefficients["nV.fit"], 
                       beta.GLM2p.r = fit.GLM2p$coefficients["r"],
                       beta.GLM2p.delta = fit.GLM2p$coefficients["delta.fit"]
  )
  dftmp2
}
```

## run simulation

```{r}
dfstat <- data.table()
nTrial.list <- c(3, 10, 25, seq(50,500,by = 50))
nEpisode <- 100

# for different numbers of trial
for (nTrial in nTrial.list) {
  
  # for each episode (different realizations of reward sequence)
  for (idxep in 1:nEpisode) {
    
    idxGroup <- 0
    if (idxep %% 100 == 1)
      cat("idxep:", idxep, "nTrial:", nTrial, format(Sys.time(), "%H:%M:%OS"), "\n")
    
    # reward sequence
    r <- c(rep(1, round(nTrial * pr)), rep(0, round(nTrial * (1-pr))))
    r <- sample(r)
    
    dfstat <- data.table()
    
    for (alpha.true in c(alpha1,alpha2)) {
      idxGroup <- idxGroup + 1
      
      for (idxSub in 1:nSub) {
        # 
        dftmp2 <- run_each_sub(alpha.true, alpha.fit, r, sigma_epsion)
        dftmp2 <- dftmp2 %>% mutate(group = as.factor(idxGroup),
                                    subject = idxSub
                                    )
        dfstat <- dplyr::bind_rows(dfstat, dftmp2)
      }
      
    }
    
    # calculate effect sizes
    es.GLM1.delta <- effsize(dfstat$beta.GLM1.delta[dfstat$group == 1],
                             dfstat$beta.GLM1.delta[dfstat$group == 2])
    
    es.GLM2.nV <- effsize(dfstat$beta.GLM2.nV[dfstat$group == 1],
                          dfstat$beta.GLM2.nV[dfstat$group == 2])
    
    # store result for each subject
    df_effectsize <- dplyr::bind_rows(df_effectsize, 
                                      data.frame(alpha1 = alpha1, alpha2 = alpha2, nTrial = nTrial, 
                                                 es.GLM1.delta = es.GLM1.delta$x12, 
                                                 es.GLM2.nV = es.GLM2.nV$x12))
  }
}
```

## Plot

```{r}
gcommon <- ggplot(df_effectsize, aes(x = nTrial)) +
  scale_colour_ptol() + 
  theme(axis.text.x = element_text(size=15),
        axis.text.y = element_text(size=15)) + 
  xlab("Number of trials") + 
  ylab("effect size (between-group)")  +
  # ylim(-1, 2.7) + 
  theme(plot.title = element_text(hjust=0.5,size=18)) 
```


```{r}
g1 <- gcommon + 
  geom_jitter(aes(y = es.GLM1.delta), width = 2, alpha = 0.3, height = 0, color = "gray30") +
      stat_function(fun=cohen.d.theory.GLM1, 
                    args = list(pr = pr, 
                                alpha.fit = alpha.fit,
                                alpha1 = alpha1, 
                           alpha2 = alpha2
                           ),
                    size = 1.5, color = "steelblue" 
      ) + 
    stat_summary(aes(y = es.GLM1.delta), fun.data=mean_se, geom="errorbar", 
               size=1.0,width=0.3,size = 2.5, color = "darkorange") + 
  stat_summary(aes(y = es.GLM1.delta), fun.y=mean, geom="point", size = 2.5, color = "darkorange") +
  labs(tag = "A") + ggtitle("GLM1, RPE")
print(g1)
```

```{r}
g2 <- gcommon + 
  geom_jitter(aes(y = es.GLM2.nV), width = 2, alpha = 0.3, height = 0, color = "gray30") +
      stat_function(fun=cohen.d.theory.GLM2.nV,
                    args = list(pr = pr, 
                                alpha.fit = alpha.fit,
                                alpha1 = alpha1,
                           alpha2 = alpha2
                           ),
                    size = 1.5, color = "steelblue" 
      ) +
    stat_summary(aes(y = es.GLM2.nV), fun.data=mean_se, geom="errorbar", 
               size=1.0,width=0.3,size = 2.5, color = "darkorange") + 
  stat_summary(aes(y = es.GLM2.nV), fun.y=mean, geom="point", size = 2.5, color = "darkorange") +
  labs(tag = "B") + ggtitle("GLM2, -V")
print(g2)
```

```{r, fig.height=4.5, fig.width=12}
gridExtra::grid.arrange(g1, g2, nrow = 1)
 
gg <- arrangeGrob(g1, g2, nrow = 1)
# ggsave(file="./figs/simuation_vs_theory_nTrial.eps", gg)
ggsave(file="./figs/simuation_vs_theory_nTrial.png", gg)
```

