---
title: "Theory (analytical results) for group comparison of regression coefficients for RW model"
author: "Kentaro Katahira"
date: "2020.09.17"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = FALSE)
```

```{r}
library(data.table)
library(tidyverse)
library(gridExtra)
library(pwr)

theme_set(theme_classic(base_size = 16))
```

# Setting

### RW model
We consider RW model
\begin{align*}
V_{t+1} &= V_t + \alpha \delta_t \\
\delta_t &= r_t - V_t
\end{align*}
where $V_1 = 0$. 

### GLMs

* GLM1
\begin{align*}
y_t = \beta_{\delta} \delta_t + \epsilon_t
\end{align*}

* GLM2
\begin{align*}
y_t =  \beta_{r} r_t + \beta_{nV} (-V) + \epsilon_t
\end{align*}

* GLM2p
\begin{align*}
y_t = \beta_{\delta} \delta_t + \beta_{r} r_t + \epsilon_t
\end{align*}
$\epsilon_t$ is Gaussian variable with mean zero, sd, $\sigma_{\epsilon} = 0.5$, 
$\beta_{\delta} = 1$, $\beta_{r} = 0$

### Setting

* 100 trials
* Reward probability: $p_r = 0.4$
* Group 1 ($N=20$, depressed): $\alpha = 0.2$
* Group 2 ($N=20$, control): $\alpha = 0.4$
* fit parameter $\hat{\alpha} = 0.3$. 


```{r}
# number of subjects for each group
nSub <- 20

# number of trials
nTrial <- 100

# reward probability
pr <- 0.4

# sd of fMRI noise
sigma_epsilon <- 0.5

# true parameter
alpha1 <- 0.2
alpha2 <- 0.4

alpha.true.list <- c(alpha1, alpha2)

# fit alpha (common to both groups)
alpha.fit <- 0.3
```

# Calculate statistical power for difference in group mean


Functions for analytical results. 
## GLM1

```{r}
beta.GLM1.delta.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.delta = 1, rho.fit = 1, rho.true = 1) {
   (2 - alpha.fit) * (alpha.true + alpha.fit)/
  (2 * (alpha.fit + alpha.true - alpha.fit * alpha.true) ) * rho.true / rho.fit 
}

beta.GLM1.delta.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 * (2-alpha.fit) / (2 * nTrial * pr * (1 - pr) * rho.fit^2)
}
```

```{r}
# group 1
M1.GLM1 <- beta.GLM1.delta.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha1)

V1.GLM1 <- beta.GLM1.delta.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha1, 
                   sigma_epsilon)

# group 2
M2.GLM1 <- beta.GLM1.delta.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha2)

V2.GLM1 <- beta.GLM1.delta.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha2, 
                   sigma_epsilon)

cohen_d.GLM1 <- (M1.GLM1 - M2.GLM1)/(sqrt((V1.GLM1 + V2.GLM1)/2))

pwr.res.GLM1 <- pwr.t.test(n = 20, sig.level = 0.05, d = cohen_d.GLM1,  
                type="t",alternative="two.sided")

cat("Power:",pwr.res.GLM1$power,"\n", "d = ", cohen_d.GLM1, 
    " M1 = ",M1.GLM1, " M2 = ",M2.GLM1,
    " V1 = ",V1.GLM1, " V2 = ",V2.GLM1, "\n")
```


## GLM2


```{r}
beta.GLM2.nV.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
   alpha.true * (2-alpha.fit) /
    (alpha.true + alpha.fit - alpha.true * alpha.fit) * beta.nV * rho.true / rho.fit 
}

beta.GLM2.nV.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 * (2 - alpha.fit) / (nTrial * alpha.fit * pr * (1 - pr) * rho.fit^2)
}

beta.GLM2.r.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
    beta.r * rho.true
}

beta.GLM2.r.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 / (nTrial * pr * (1 - pr))
}
```

negative value (-V)

```{r}
# group 1
M1.GLM2 <- beta.GLM2.nV.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha1)

V1.GLM2 <- beta.GLM2.nV.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha1, 
                   sigma_epsilon)

# group 2
M2.GLM2 <- beta.GLM2.nV.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha2)

V2.GLM2 <- beta.GLM2.nV.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha2, 
                   sigma_epsilon)

cohen_d <- (M1.GLM2 - M2.GLM2)/(sqrt((V1.GLM2 + V2.GLM2)/2))

pwr.res <- pwr.t.test(n = 20, sig.level = 0.05, d = cohen_d, 
                type="t",alternative="two.sided")

cat("Power:",pwr.res$power,"\n", "d = ", cohen_d, 
    " M1 = ", M1.GLM2, " M2 = ", M2.GLM2,
    " V1 = ", V1.GLM2, " V2 = ", V2.GLM2, "\n")
```

reward (r)

```{r}
# group 1
M1.GLM2 <- beta.GLM2.r.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha1)

V1.GLM2 <- beta.GLM2.r.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha1, 
                   sigma_epsilon)

# group 2
M2.GLM2 <- beta.GLM2.r.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha2)

V2.GLM2 <- beta.GLM2.r.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha2, 
                   sigma_epsilon)

cohen_d <- (M1.GLM2 - M2.GLM2)/(sqrt((V1.GLM2 + V2.GLM2)/2))

pwr.res <- pwr.t.test(n = 20, sig.level = 0.05, d = cohen_d, 
                type="t",alternative="two.sided")

cat("Power:",pwr.res$power,"\n", "d = ", cohen_d, 
    " M1 = ", M1.GLM2, " M2 = ", M2.GLM2,
    " V1 = ", V1.GLM2, " V2 = ", V2.GLM2, "\n")
```



## GLM2p

Function for analytical results. 

```{r}
beta.GLM2p.delta.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.delta = 1, rho.fit = 1, rho.true = 1) {
   alpha.true * (2-alpha.fit) /
    (alpha.true + alpha.fit - alpha.true * alpha.fit) * beta.delta * rho.true / rho.fit 
}

beta.GLM2p.delta.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.delta = 1, rho.fit = 1, rho.true = 1) {
  sigma_epsilon^2 * (2 - alpha.fit) / (nTrial * alpha.fit * pr * (1 - pr) * rho.fit^2)
}

beta.GLM2p.r.mean <- function(nTrial, pr, alpha.fit, alpha.true, beta.delta = 1, beta.r = 0, rho.fit = 1, rho.true = 1) {
    (alpha.fit - alpha.true)/(alpha.fit + alpha.true - alpha.fit * alpha.true) * beta.delta * rho.true / rho.fit + beta.r
}

beta.GLM2p.r.var <- function(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.delta = 1, rho.fit = 1, rho.true = 1) {
  2 * sigma_epsilon^2 / (nTrial * alpha.fit * pr * (1 - pr))
}
```


```{r}
# group 1
M1.GLM2p <- beta.GLM2p.delta.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha1)

V1.GLM2p <- beta.GLM2p.delta.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha1, 
                   sigma_epsilon)

# group 2
M2.GLM2p <- beta.GLM2p.delta.mean(nTrial, pr, 
                     alpha.fit = alpha.fit, 
                     alpha.true = alpha2)

V2.GLM2p <- beta.GLM2p.delta.var(nTrial, pr, 
                   alpha.fit = alpha.fit, 
                   alpha.true = alpha2, 
                   sigma_epsilon)

cohen_d <- (M1.GLM2p - M2.GLM2p)/(sqrt((V1.GLM2p + V2.GLM2p)/2))

pwr.res <- pwr.t.test(n = 20, sig.level = 0.05, d = cohen_d, 
                type="t",alternative="two.sided")

cat("Power:",pwr.res$power,"\n", "d = ", cohen_d, 
    " M1 = ", M1.GLM2p, " M2 = ", M2.GLM2p,
    " V1 = ", V1.GLM2p, " V2 = ", V2.GLM2p, "\n")
```



# Plot analytical results of mean regression coefficients


## common ggplot setting

```{r}
gcommon <- ggplot(data.frame(alpha.fit = c(0.1, 0.9)), aes(x = alpha.fit)) +
  scale_color_manual(values =c("indianred","steelblue")) + 
  labs(color="True value:") +
  geom_vline(xintercept = alpha1, colour = "gray44", linetype ="dashed") +
  geom_vline(xintercept = alpha2, colour = "gray44", linetype ="dashed") +
  geom_vline(xintercept = alpha.fit, colour = "gray44") +
  theme(legend.position = "none", aspect.ratio=1) + 
  theme(plot.title = element_text(hjust=0.5,size=18)) + 
    labs(x = TeX('fit learning rate, $\\hat{\\alpha}$'),
        y =TeX('expected beta value, $E\\[\\hat{\\beta}\\]$')) +
  ylim(0, 1.8)
```



## GLM1 

```{r}
g.m.GLM1 <- gcommon + 
  geom_hline(yintercept = 1, colour = "gray44", linetype ="dashed") +
  mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM1.delta.mean, 
                    args = list(nTrial = nTrial, 
                                pr = pr, 
                                alpha.true = alpha.true),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  labs( title = "GLM1 (RPE only)",
        tag = "A")
```



## GLM2
```{r}
g.m.GLM2 <- gcommon + 
  geom_hline(yintercept = 1, colour = "gray44", linetype ="dashed") +
  mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM2.nV.mean, 
                    args = list(nTrial = nTrial, 
                                pr = pr, 
                                alpha.true = alpha.true),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
    mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM2.r.mean, 
                    args = list(nTrial = nTrial, 
                                pr = pr, 
                                alpha.true = alpha.true),
                    size = 1, linetype = "dashed", 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  labs(title = "GLM2 (reward + (-V))", tag = "B")
```



```{r, fig.height=4.5, fig.width=11}
gridExtra::grid.arrange(g.m.GLM1, #g.m.GLM2p, 
                        g.m.GLM2, nrow = 1)
```



# SD of regression coefficients

Define functions for calculating SD
```{r}
# GLM1
beta.GLM1.delta.sd <- function(nTrial, pr, alpha.fit = alpha.fit, alpha.true, 
                               sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  sqrt(beta.GLM1.delta.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit, rho.true))
}

# GLM2p
#  r
beta.GLM2p.r.sd <- function(nTrial, pr, alpha.fit, alpha.true, 
                           sigma_epsilon = sigma_epsilon, beta.delta = 1, rho.fit = 1, rho.true = 1) {
  sqrt(beta.GLM2p.r.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.delta, rho.fit, rho.true))
}
#  delta
beta.GLM2p.delta.sd <- function(nTrial, pr, alpha.fit, alpha.true, 
                               sigma_epsilon = sigma_epsilon, beta.delta = 1, rho.fit = 1, rho.true = 1) {
  sqrt(beta.GLM2p.delta.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.delta, rho.fit, rho.true))
}

# GLM2
#  r
beta.GLM2.r.sd <- function(nTrial, pr, alpha.fit, alpha.true, 
                           sigma_epsilon = sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sqrt(beta.GLM2.r.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r, beta.nV, rho.fit, rho.true))
}
#  nV
beta.GLM2.nV.sd <- function(nTrial, pr, alpha.fit, alpha.true, 
                            sigma_epsilon = sigma_epsilon, beta.r = 1, beta.nV = 1, rho.fit = 1, rho.true = 1) {
  sqrt(beta.GLM2.nV.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r, beta.nV, rho.fit, rho.true))
}
```


```{r}
gcommon <- ggplot(data.frame(alpha.fit = c(0.1, 0.9)), aes(x = alpha.fit)) + 
  scale_color_manual(values =c("indianred","steelblue")) + 
  labs(color="True value:") +
  geom_vline(xintercept = alpha1, colour = "gray44", linetype ="dashed") +
  geom_vline(xintercept = alpha2, colour = "gray44", linetype ="dashed") +
  geom_vline(xintercept = alpha.fit, colour = "gray44") +
  theme(legend.position = "none", aspect.ratio=1) + 
  theme(plot.title = element_text(hjust=0.5,size=18)) + 
    labs(x = TeX('fit learning rate, $\\hat{\\alpha}$'),
        y = 'S.D. of beta value') +
  ylim(0, 1.8)
```


## GLM1

```{r}
g.sd.GLM1 <- gcommon + 
  mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM1.delta.sd, 
                    args = list(nTrial = nTrial, 
                                pr = pr, sigma_epsilon = sigma_epsilon, 
                                alpha.true = alpha.true),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + ylim(0,0.6) + 
  labs(tag = "C") 
```


## GLM2

```{r}
g.sd.GLM2 <- gcommon + 
    mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM2.r.sd, 
                    args = list(nTrial = nTrial, 
                                pr = pr, sigma_epsilon = sigma_epsilon, 
                                alpha.true = alpha.true),
                    size = 1, linetype = "dashed", 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  mapply(
    function(alpha.true, co) 
      stat_function(fun=beta.GLM2.nV.sd, 
                    args = list(nTrial = nTrial, 
                                pr = pr, sigma_epsilon = sigma_epsilon, 
                                alpha.true = alpha.true),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) +  ylim(0,0.6) + 
  labs(tag = "D") 
```

```{r fig.height=4.5, fig.width=7}
gridExtra::grid.arrange(g.sd.GLM1, # g.sd.GLM2p, 
                        g.sd.GLM2, nrow = 1)
# 
# gg <- arrangeGrob(g.sd.GLM1, # g.sd.GLM2p, 
#                   g.sd.GLM2, nrow = 1)
# ggsave(file="./figs/theory_sd_alpha.eps", gg)
```

# Effect size (single sample)

Define functions for calculating effect size
```{r}
# GLM1
effectsize.GLM1.delta <- function(nTrial, pr, alpha.fit, alpha.true, 
                                  beta.delta = 1, sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  beta.GLM1.delta.mean(nTrial, pr, alpha.fit, alpha.true, beta.delta, rho.fit, rho.true)/
    sqrt(beta.GLM1.delta.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit, rho.true))
}

# GLM2p
effectsize.GLM2p.delta <- function(nTrial, pr, alpha.fit, alpha.true, 
                                  beta.delta = 1, beta.r = 0, sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  beta.GLM2p.delta.mean(nTrial, pr, alpha.fit, alpha.true, beta.delta, rho.fit, rho.true)/
    sqrt(beta.GLM2p.delta.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit, rho.true))
}

effectsize.GLM2p.r <- function(nTrial, pr, alpha.fit, alpha.true, 
                                  beta.delta = 1, beta.r = 0, sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  beta.GLM2p.r.mean(nTrial, pr, alpha.fit, alpha.true, beta.delta, beta.r, rho.fit, rho.true)/
    sqrt(beta.GLM2p.r.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, rho.fit, rho.true))
}

# GLM2
effectsize.GLM2.nV <- function(nTrial, pr, alpha.fit, alpha.true, 
                                  beta.r= 1, beta.nv = 1, sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  beta.GLM2.nV.mean(nTrial, pr, alpha.fit, alpha.true, beta.r, beta.nv, rho.fit, rho.true)/
    sqrt(beta.GLM2.nV.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r, beta.nv, rho.fit, rho.true))
}

effectsize.GLM2.r <- function(nTrial, pr, alpha.fit, alpha.true, 
                                  beta.r = 1, beta.nv = 1, sigma_epsilon = sigma_epsilon, rho.fit = 1, rho.true = 1) {
  beta.GLM2.r.mean(nTrial, pr, alpha.fit, alpha.true, beta.r, beta.nv, rho.fit, rho.true)/
    sqrt(beta.GLM2.r.var(nTrial, pr, alpha.fit, alpha.true, sigma_epsilon, beta.r, beta.nv, rho.fit, rho.true))
}
```

## GLM1

```{r}
g.effect.GLM1 <- gcommon + 
  mapply(
    function(alpha.true, co) 
      stat_function(fun=effectsize.GLM1.delta, 
                    args = list(nTrial = nTrial, 
                                pr = pr, sigma_epsilon = sigma_epsilon, 
                                alpha.true = alpha.true),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  ylab("effect size (single group)") + labs(tag = "E") + ylim(0,13)  
```

## GLM2

```{r}
g.effect.GLM2 <- gcommon + 
  mapply(
    function(alpha.true, co) 
      stat_function(fun=effectsize.GLM2.nV, 
                    args = list(nTrial = nTrial, 
                                pr = pr, 
                                alpha.true = alpha.true,
                                sigma_epsilon = sigma_epsilon),
                    size = 1, 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  mapply(
    function(alpha.true, co) 
      stat_function(fun=effectsize.GLM2.r, 
                    args = list(nTrial = nTrial, 
                                pr = pr, 
                                alpha.true = alpha.true,
                                sigma_epsilon = sigma_epsilon),
                    size = 1, linetype = "dashed", 
                    aes_q(color=co)
      ), alpha.true.list, sprintf("alpha = %.2f", alpha.true.list)
  ) + 
  labs(tag = "F") + ylim(0,13) + ylab("effect size (single group)")
```

```{r, fig.height=4.5, fig.width=7}
gridExtra::grid.arrange(g.effect.GLM1, 
                        g.effect.GLM2, nrow = 1)
```


```{r, fig.height=12, fig.width=7}
gg <- arrangeGrob(g.m.GLM1, g.m.GLM2, 
                  g.sd.GLM1, g.sd.GLM2,
                  g.effect.GLM1, g.effect.GLM2, 
                  nrow = 3)
plot(gg)
ggsave(file="./figs/theory_alpha.eps", gg)
```
